---
title: "Final Biology 201 Project"
author: "Seth Woodbury, Julie Heng, Maira Gajda, Karan Nayak"
date: "11/18/2019"
output: html_document
---

Github Link: https://github.com/hengj24/bio201_project


# Research Question:

Do any supplements cause a significant mean decrease in human gut pH from week 1 (pre-supplement) to week 3 (post-supplement)?

Hypothesis: We hypothesize that frequency is important when consuming resistant starch and that some supplements will have more significant results than others because they are not all made up of the same thing (which means different food sources for microbes).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/bio201_project")
```

### Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq); packageVersion("phyloseq")
set.seed(7)
```

## Import Data
```{r}
#Import Data
shared_wkly <- read_delim(file = "raw_data/shared_wkly.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


biographical <- read_delim(file = "raw_data/biographical.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


ph_wkly <- read_delim(file = "raw_data/DB_v_0.08/pH_wkly.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


#Look at pH Data
ph_wkly
count(ph_wkly, Supplement_consumed)
count(ph_wkly, Frequency)
```
There are 6 supplements:

1. Banana (n=32)

2. BRMPS (n=365)

3. HiMaize+BRMPS (n=93)

4. LOODAT (n=35)

5. none (n=73)

6. transition_HiMaize (n=87)

and 2 with NA (remove)


## Curate pH Weekly Data
```{r}
#Rename to our standards and filter data
curated_ph_wkly <- ph_wkly %>%
  rename_all(tolower) %>%
  filter(quantity_compliant == "yes" | quantity_compliant == "none",
         use_data == "yes",
         study_week == "week1" | study_week == "week3",
         frequency != "NA") 
write_delim(curated_ph_wkly, path = "curated_data/curated_ph_wkly.txt", delim = "\t")
curated_ph_wkly
```

Now we have a common standard for all of our data. From here, we can further subset for specific conditions, but this data set contains the general conditions for all further calculations.


# Supplement Specific Assumption & Statistical Testing with Plots

It is important to note that there was no pH data for participants who consumed Bananas for week 1, and thus we could not compute a statistical analysis on it.

## 2x Daily BRMPS
```{r} 
#2x Daily BRMPS


#Further Curation
brmps_wk1_2x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 


brmps_wk3_2x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 


#Normality
shapiro.test(brmps_wk1_2x$ph_mean) #p-value = .3826
shapiro.test(brmps_wk3_2x$ph_mean) #p-value = .006713


#Normality Plots
ggplot(brmps_wk1_2x, aes(x = ph_mean)) + geom_histogram()
ggplot(brmps_wk3_2x, aes(x = ph_mean)) + geom_histogram()

qqnorm(brmps_wk1_2x$ph_mean); qqline(brmps_wk1_2x$ph_mean)
qqnorm(brmps_wk3_2x$ph_mean); qqline(brmps_wk3_2x$ph_mean)


#Final Joined Data
brmps_2x <- inner_join(x = brmps_wk1_2x, y = brmps_wk3_2x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(brmps_ph_mean_2x_wk1 = ph_mean.x,
         brmps_ph_mean_2x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(brmps_2x, path = "curated_data/brmps_2x.txt", delim = "\t")
brmps_2x

#Sample Size + Varience Assumptions
str(brmps_2x) # n=27
var.test(x = brmps_2x$brmps_ph_mean_2x_wk1, 
         y = brmps_2x$brmps_ph_mean_2x_wk3, 
         alternative = "greater") #p-value = .9512


#Statistical Tests 
wilcox.test(x = brmps_2x$brmps_ph_mean_2x_wk1, 
            y = brmps_2x$brmps_ph_mean_2x_wk3, 
            alternative = "greater", paired = TRUE, 
            var.equal = TRUE)  #p-value = 0.0002761


#BRMPS 2x Daily Plot
brmps2_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         study_week == "week1" | study_week == "week3",
         frequency == "2xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS Supplement (2x daily)") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = brmps2_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


brmps2_plot
```

We have an extremely low p-value (less than alpha = .001) indicating that BRMPS taken 2x daily did lower pH in participants from week 1 to week 3 on average. In calculating the conclusion, we decided to use a non-parametric t-test (wilcox) because week 3 had non-normal data and week 1 did not follow the Normal Q-Q plot very accurately and the histogram had a slight bimodal look. Additionally because the sample size was less than 30 we decided to be safe and use the non-parametric test; variences were equal though for this test.


## 2x Daily Transition Maize
```{r}
#2x Daily Transition Maize


#Further Curation
transition_wk1_2x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "transition_HiMaize", 
         frequency == "2xdaily") 


transition_wk3_2x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "transition_HiMaize", 
         frequency == "2xdaily") 


#Normaily + Varience Assumptions
shapiro.test(transition_wk1_2x$ph_mean) #p-value = .8910
shapiro.test(transition_wk3_2x$ph_mean) #p-value = .5083


#Normality Plots
ggplot(transition_wk1_2x, aes(x = ph_mean)) + geom_histogram()
ggplot(transition_wk3_2x, aes(x = ph_mean)) + geom_histogram()

qqnorm(transition_wk1_2x$ph_mean); qqline(transition_wk1_2x$ph_mean)
qqnorm(transition_wk3_2x$ph_mean); qqline(transition_wk3_2x$ph_mean)


#Final Joined Data
transition_2x <- inner_join(x = transition_wk1_2x, y = transition_wk3_2x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(transition_ph_mean_2x_wk1 = ph_mean.x,
         transition_ph_mean_2x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(transition_2x, path = "curated_data/transition_2x.txt", delim = "\t")
transition_2x


#Sample Size + Varience Assumptions
str(transition_2x) # n=21
var.test(x = transition_2x$transition_ph_mean_2x_wk1, 
         y = transition_2x$transition_ph_mean_2x_wk3, 
         alternative = "greater") #p-value = .9395


#Statistical Tests
t.test(x = transition_2x$transition_ph_mean_2x_wk1, y = transition_2x$transition_ph_mean_2x_wk3, 
       alternative = "greater", paired = TRUE, 
       var.equal = TRUE) #p-value = 0.007215


#Transition HiMaize 2x Daily Plot
transition_himaize_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "transition_HiMaize",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week)) +
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after 2x Transition HiMaize") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "transition_himaize_plot.pdf",
          plot = transition_himaize_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


transition_himaize_plot
```

After running significance testing for Transition Maize taken 2x daily we found that there was a signifiant p-value (less than alpha = .01) providing evidence to reject the null hypothesis and conclude that this supplement did decrease the pH in participants from week 1 to week 3 on average. The sample size was not ideal, but both shapiro tests indicated that each subsetted data set was approximately normal which we believed was sufficient evidence to carry out a parametric test (t-test); the variences were equal for the data as well.


## 1x Daily BRMPS
```{r}
#1x Daily BRMPS 


#Further Curation
brmps_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 


brmps_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 


#Normaily + Varience Assumptions
shapiro.test(brmps_wk1_1x$ph_mean) #p-value = 0.01669
shapiro.test(brmps_wk3_1x$ph_mean) #p-value = .08154


#Normality Plots
ggplot(brmps_wk1_1x, aes(x = ph_mean)) + geom_histogram()
ggplot(brmps_wk3_1x, aes(x = ph_mean)) + geom_histogram()

qqnorm(brmps_wk1_1x$ph_mean); qqline(brmps_wk1_1x$ph_mean)
qqnorm(brmps_wk3_1x$ph_mean); qqline(brmps_wk3_1x$ph_mean)


#Final Joined Data
brmps_1x <- inner_join(x = brmps_wk1_1x, y = brmps_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(brmps_ph_mean_1x_wk1 = ph_mean.x,
         brmps_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(brmps_1x, path = "curated_data/brmps_1x.txt", delim = "\t")
brmps_1x


#Sample Size + Varience Assumptions
str(brmps_1x) # n=62
var.test(x = brmps_1x$brmps_ph_mean_1x_wk1, 
         y = brmps_1x$brmps_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .6063


#Statistical Tests 
t.test(x = brmps_1x$brmps_ph_mean_1x_wk1, y = brmps_1x$brmps_ph_mean_1x_wk3, 
       alternative = "greater", paired = TRUE, 
       var.equal = TRUE) #p-value = 0.1133


#BRMPS 1x Daily Plot
brmps1_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS Supplement (1x daily)") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = brmps1_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


brmps1_plot
```

Although close to our alpha value of .10, we still could not reject our null hypothesis because the p-value was higher than the alpha value. Thus, BRMPS consumed 1x daily did not appear to significantly decrease the pH in participants' guts on average from week 1 to week 3. We decided that a parametric test would be sufficient since the qq plots showed normal data and the histograms appeared to be approximately normal, even thought the sample size was not greater than 30. Additionally, we found the variences to be equal between the subsetted data sets.


## 1x Daily HiMaize+BRMPS
```{r}
#1x Daily HiMaize+BRMPS


#Further Curation
himaize_brmps_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "HiMaize+BRMPS", 
         frequency == "1xdaily") 


himaize_brmps_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "HiMaize+BRMPS", 
         frequency == "1xdaily") 


#Normaily + Varience Assumptions
shapiro.test(himaize_brmps_wk1_1x$ph_mean) #p-value = .2955
shapiro.test(himaize_brmps_wk3_1x$ph_mean) #p-value = .05622


#Normality Plots
ggplot(himaize_brmps_wk1_1x, aes(x = ph_mean)) + geom_histogram()
ggplot(himaize_brmps_wk3_1x, aes(x = ph_mean)) + geom_histogram()

qqnorm(himaize_brmps_wk1_1x$ph_mean); qqline(himaize_brmps_wk1_1x$ph_mean)
qqnorm(himaize_brmps_wk3_1x$ph_mean); qqline(himaize_brmps_wk3_1x$ph_mean)


#Final Joined Data
himaize_brmps_1x <- inner_join(x = himaize_brmps_wk1_1x, y = himaize_brmps_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(himaize_brmps_ph_mean_1x_wk1 = ph_mean.x,
         himaize_brmps_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(himaize_brmps_1x, path = "curated_data/himaize_brmps_1x.txt", delim = "\t")
himaize_brmps_1x


#Sample Size + Varience Assumptions
str(himaize_brmps_1x) # n=23
var.test(x = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk1, 
         y = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .489


#Statistical Tests 
wilcox.test(x = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk1, 
            y = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk3, 
            alternative = "greater", paired = TRUE, 
            var.equal = TRUE) #p-value = 0.05289


#HiMaize+BRMPS 1x Daily Plot
himaize_brmps_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "HiMaize+BRMPS",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after HiMaize+BRMPS Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = himaize_brmps_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


himaize_brmps_plot
```

Since we are looking at complex and diverse biological systems, we thought it was reasonable to use an alpha value of .10. Thus, in this case we found a significant p-value that was less than alpha = .10 providing us with sufficient evidence to reject the null hypothesis and accept the alternative concluding that consuming HiMaize+BRMPS supplement 1x daily did decrease pH on average in individuals from week 1 to week 3. We used a non-parametric test (wilcox) with equal variences because week 3 had non-normal data according to the shapiro test and the graph for week 1 appeared to be bimodal in addition to a low sample size.


## 1x Daily LOODAT
```{r}
#1x Daily LOODAT


#Further Curation
loodat_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "LOODAT", 
         frequency == "1xdaily") 


loodat_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "LOODAT", 
         frequency == "1xdaily") 


#Normaily + Varience Assumptions
shapiro.test(loodat_wk1_1x$ph_mean) #p-value = .1888
shapiro.test(loodat_wk3_1x$ph_mean) #p-value = .0213


#Normality Plots
ggplot(loodat_wk1_1x, aes(x = ph_mean)) + geom_histogram()
ggplot(loodat_wk3_1x, aes(x = ph_mean)) + geom_histogram()

qqnorm(loodat_wk1_1x$ph_mean); qqline(loodat_wk1_1x$ph_mean)
qqnorm(loodat_wk3_1x$ph_mean); qqline(loodat_wk3_1x$ph_mean)


#Final Joined Data
loodat_1x <- inner_join(x = loodat_wk1_1x, y = loodat_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(loodat_ph_mean_1x_wk1 = ph_mean.x,
         loodat_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(loodat_1x, path = "curated_data/loodat_1x.txt", delim = "\t")
loodat_1x


#Sample Size + Varience Assumptions
str(loodat_1x) # n=15
var.test(x = loodat_1x$loodat_ph_mean_1x_wk1, 
         y = loodat_1x$loodat_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .3212


#Statistical Tests 
wilcox.test(x = loodat_1x$loodat_ph_mean_1x_wk1, 
            y = loodat_1x$loodat_ph_mean_1x_wk3, 
            alternative = "greater", paired = TRUE, 
            var.equal = TRUE) #p-value = 0.9024


#LOODAT 1x Daily Plot
loodat_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "LOODAT",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after LOODAT Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = loodat_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


loodat_plot
```

The p-value here was far from significant, which would have required it to be lower than alpha = .10. Since it was above this value, we cannot reject the null hypothesis and therefore must conclude that consuming LOODAT 1x daily did not decrease the pH on average in participants' guts from week 1 to week 3. We used a non-parametric test (wilcox) with equal variences because both subsetted data sets had bimodal data and there was a low sample size.


## No Supplement Consumed
```{r}
#No Supplement Consumed 


#Further Curation
no_supplement_wk1 <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "none", 
         frequency == "-") 


no_supplement_wk3 <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "none", 
         frequency == "-") 


#Normaily + Varience Assumptions
shapiro.test(no_supplement_wk1$ph_mean) #p-value = .1029
shapiro.test(no_supplement_wk3$ph_mean) #p-value = .03968


#Normality Plots
ggplot(no_supplement_wk1, aes(x = ph_mean)) + geom_histogram()
ggplot(no_supplement_wk3, aes(x = ph_mean)) + geom_histogram()

qqnorm(no_supplement_wk1$ph_mean); qqline(no_supplement_wk1$ph_mean)
qqnorm(no_supplement_wk3$ph_mean); qqline(no_supplement_wk3$ph_mean)


#Final Joined Data
no_supplement <- inner_join(x = no_supplement_wk1, y = no_supplement_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(no_supplement_ph_mean_wk1 = ph_mean.x,
         no_supplement_ph_mean_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
write_delim(no_supplement, path = "curated_data/no_supplement.txt", delim = "\t")
no_supplement

#Sample Size + Varience Assumptions
str(no_supplement) # n=24
var.test(x = no_supplement$no_supplement_ph_mean_wk1, 
         y = no_supplement$no_supplement_ph_mean_wk3, 
         alternative = "greater") #p-value = 0.2721


#Statistical Tests 
wilcox.test(x = no_supplement$no_supplement_ph_mean_wk1, 
            y = no_supplement$no_supplement_ph_mean_wk3, 
            alternative = "greater", paired = TRUE, 
            var.equal = TRUE) #p-value = 0.7781


#No Supplement Plot
no_supplement_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "none",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH without Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = no_supplement_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


no_supplement_plot
```

In testing our control (no supplement) group, we hypothesized that there would be no significant result, and that is exactly what we got as the p-value was greater than alpha = .10. Thus we accept the null hypothesis stating that there was not a decrease on average for participants who did not consume a supplement from week 1 to week 3. We surprisingly found the data sets to be bimodal when we graphed them and did we not have a large sample size leading us to use a non-parametric (wilcox) statistical test. In addition, we found the variences to be equal between the two subsetted data sets.


## Finding the Mean Difference in pH between Week 3 and Week 1 for Each Supplement (Week 3 - Week 1)

The fact that some supplements caused a significant decrease in pH and others did not suggests that not all supplements produce the same results and that there might be an overarching supplement that caused the lowest drop in pH. Our goal here is to determine which supplement causes the largest mean difference in pH. We will only test the statistically significant data sets found above:
```{r}
#BRMPS 2x Daily Mean pH Difference
mean_diff_brmps_2x <- brmps_2x %>%
  select(participant_id, brmps_ph_mean_2x_wk1, brmps_ph_mean_2x_wk3) %>%
  mutate(ph_difference = brmps_ph_mean_2x_wk3 - brmps_ph_mean_2x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

mean_diff_brmps_2x #pH Difference = -0.3414815
```

This calculation tells us that participants who consumed BRMPS 2x daily had an average decrease of .34 in pH from week 1 to week 3.


```{r}
#Transition Maize 2x Daily Mean pH Difference
mean_diff_transition_2x <- transition_2x %>%
  select(participant_id, transition_ph_mean_2x_wk1, transition_ph_mean_2x_wk3) %>%
  mutate(ph_difference = transition_ph_mean_2x_wk3 - transition_ph_mean_2x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

mean_diff_transition_2x #pH Difference = -0.2895238
```

This calculation tells us that participants who consumed Transition Maize 2x daily had an average decrease of .29 in pH from week 1 to week 3.


```{r}
#HiMaize+BRMPS 1x Daily Mean pH Difference
mean_diff_himaize_brmps_1x <- himaize_brmps_1x %>%
  select(participant_id, himaize_brmps_ph_mean_1x_wk1, himaize_brmps_ph_mean_1x_wk3) %>%
  mutate(ph_difference = himaize_brmps_ph_mean_1x_wk3 - himaize_brmps_ph_mean_1x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

mean_diff_himaize_brmps_1x #pH Difference = -0.2556522
```

This calculation tells us that participants who consumed HiMaize+BRMPS 1x daily had an average decrease of .26 in pH from week 1 to week 3.

As we can see, 2x Daily BRMPS causes the largest mean decrease in pH from weak 1 to weak 3, and thus we will focus on this supplement for the remainder of our statistical study.


## "Lacto" Genre Bacteria and pH Correlations in BRMPS 2x Daily Individuals

Hypothesis: We hypothesize that there will be a negative linear correlation between pH and Lacto Bacteria because Lacto Bacteria produce acidic biproducts that would lower the pH.
```{r}
#Import Data
seq_var_table <- read_delim("raw_data/species_avg_shared.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  rename_all(tolower) %>%
  select(participant_id, study_week, semester, starts_with("Lactoc"), starts_with("Lactob"))
seq_var_table
```


```{r}
#Lactobacillus


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacillus <- inner_join(seq_var_table, brmps_2x_phylo) %>% #combine two phyloseq objects created above
  select(-starts_with("lactoco")) #get rid of lactococcus
lactobacillus


#Calculate Row Sums
lactobacillus_total <-  rowSums(lactobacillus[,4:39]) 
lactobacillus_total


#Mutate Row Sums into Previous Table
new_lactobacillus <- lactobacillus %>% 
  mutate(lactobacillus = lactobacillus_total) %>%
  select(participant_id, study_week, semester, lactobacillus, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacillus


#Correlation Data With Graph
wk_3_lactobacillus <- new_lactobacillus %>%
  filter(study_week == "week3")
write_delim(wk_3_lactobacillus, path = "curated_data/wk_3_lactobacillus.txt", delim = "\t")


wk_3_lactobacillus_plot <- wk_3_lactobacillus %>%
  ggplot(aes(x = ph_mean,
             y = lactobacillus)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactobacillus Bacteria") 
wk_3_lactobacillus_plot


#Correlation Test
lactobacillus_correlation <- wk_3_lactobacillus %>%
  lm(ph_mean ~ lactobacillus, data = .) #test relationship
summary(lactobacillus_correlation) #view results 
lactobacillus_correlation

#p-value = .4869
#R-squared = .01953
```

Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .4869) indicates that the null hypothesis is true meaning there is no correlation between pH and Lactobacillus genre bacteria. Additionally, the small r-squared value indicates that only about 1.9% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis which is opposite to what we hypothesized, as here the number of Lactobacillus bacteria increases as pH increases.


```{r}
#Lactococcus 


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactococcus <- inner_join(seq_var_table, brmps_2x_phylo) %>% #combine two phyloseq objects created above
  select(-starts_with("lactoba")) #get rid of lactococcus
lactococcus


#Calculate Row Sums
lactococcus_total <-  rowSums(lactococcus[,4:8]) 
lactococcus_total


#Mutate Row Sums into Previous Table
new_lactococcus <- lactococcus %>% 
  mutate(lactococcus = lactococcus_total) %>%
  select(participant_id, study_week, semester, lactococcus, ph_median, ph_mean, frequency, supplement_consumed)
new_lactococcus


#Correlation Data With Graph
wk_3_lactococcus <- new_lactococcus %>%
  filter(study_week == "week3")
write_delim(wk_3_lactococcus, path = "curated_data/wk_3_lactococcus.txt", delim = "\t")


wk_3_lactococcus_plot <- wk_3_lactococcus %>%
  ggplot(aes(x = ph_mean,
             y = lactococcus)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactococcus Bacteria") 
wk_3_lactococcus_plot


#Correlation Test
lactococcus_correlation <- wk_3_lactococcus %>%
  lm(ph_mean ~ lactococcus, data = .) #test relationship
summary(lactococcus_correlation) #view results 
lactococcus_correlation

#p-value = .3089
#R-squared = .04137
```

Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .3089) indicates that the null hypothesis is true meaning there is no correlation between pH and Lactococcus genre bacteria. Additionally, the small r-squared value indicates that only about 4.1% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis, which is opposite to what we hypothesized, as here the number of Lactococcus bacteria increases as pH increases.


We figured since everybody has a different microbiome, that some individuals may contain Lactococcus or just Lactobacillus genre bacteria and since they perform similar functions we thought it would be reasonable to combine them into one data set and see if there was a correlation between "Lacto" Bacteria and pH to compensate for different microbiome makeups.
```{r}
#"Lacto" Genre Bacteria 


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacteria <- inner_join(seq_var_table, brmps_2x_phylo) 
lactobacteria


#Calculate Row Sums
lactobacteria_total <-  rowSums(lactobacteria[,4:44]) 
lactobacteria_total


#Mutate Row Sums into Previous Table
new_lactobacteria <- lactobacteria %>% 
  mutate(lactobacteria = lactobacteria_total) %>%
  select(participant_id, study_week, semester, lactobacteria, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacteria


#Correlation Data With Graph
wk_3_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week3")
write_delim(new_lactobacteria, path = "curated_data/new_lactobacteria.txt", delim = "\t")


wk_3_lactobacteria_plot <- wk_3_lactobacteria %>%
  ggplot(aes(x = ph_mean,
             y = lactobacteria)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactate Producers") 
wk_3_lactobacteria_plot


#Correlation Test
lactobacteria_correlation <- wk_3_lactobacteria %>%
  lm(ph_mean ~ lactobacteria, data = .) #test relationship
summary(lactobacteria_correlation) #view results 
lactobacteria_correlation

#p-value = .4622
#R-squared = .02182
```

Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .4622) indicates that the null hypothesis is true meaning there is no correlation between pH and "Lacto" bacteria. Additionally, the small r-squared value indicates that only about 2.2% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis, which is opposite to what we hypothesized, as here the number of "Lacto" bacteria increases as pH increases.


## Did the Number of Lactate Producers Rise?

Hypothesis: We hypothesize that the number of lactate producers will rise because the resistant starch would act as a food source to help fuel the energy demands of cellular growth and reproduction.

Because our correlation tests did not find any significant results, we decided to look a little deeper to see if the supplement BRMPS taken 2x daily did at least increase the number of Lactate producers (Lactococcus and Lactobacillus bacteria). These Lactate producers may not be correlated with more acidic conditions, but they confer a variety of health benefits such as decreased gut permeability (inhibiting toxic bacterial biproducts from being absorbed) as well as disease resistance and increased immunity. Additionally, they produce acidic biproducts which may contribute to lower gut pH.
```{r}
#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacteria <- inner_join(seq_var_table, brmps_2x_phylo) 
lactobacteria


#Calculate Row Sums
lactobacteria_total <-  rowSums(lactobacteria[,4:44]) 
lactobacteria_total


#Mutate Row Sums into Previous Table
new_lactobacteria <- lactobacteria %>% 
  mutate(lactobacteria = lactobacteria_total) %>%
  select(participant_id, study_week, semester, lactobacteria, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacteria


#Weekly Data 
wk_1_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week1")


wk_3_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week3")


#Normality
shapiro.test(wk_1_lactobacteria$lactobacteria) #p-value = 6.575e-08
shapiro.test(wk_3_lactobacteria$lactobacteria) #p-value = 2.305e-07


#Normality Plots
ggplot(wk_1_lactobacteria, aes(x = lactobacteria)) + geom_histogram()
ggplot(wk_3_lactobacteria, aes(x = lactobacteria)) + geom_histogram()

qqnorm(wk_1_lactobacteria$lactobacteria); qqline(wk_1_lactobacteria$lactobacteria)
qqnorm(wk_3_lactobacteria$lactobacteria); qqline(wk_3_lactobacteria$lactobacteria)


#Final Joined Data
stat_lactobacteria <- inner_join(x = wk_1_lactobacteria, y = wk_3_lactobacteria,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed")) %>%
  rename(lactobacteria_wk1_abund = lactobacteria.x,
         lactobacteria_wk3_abund = lactobacteria.y) %>%
  select(-starts_with("study_week"))
write_delim(stat_lactobacteria, path = "curated_data/stat_lactobacteria.txt", delim = "\t")
stat_lactobacteria


#Sample Size + Varience Assumptions
str(stat_lactobacteria) # n=26
var.test(x = stat_lactobacteria$lactobacteria_wk1_abund, 
         y = stat_lactobacteria$lactobacteria_wk3_abund, 
         alternative = "greater") #p-value= 1


#Statistical Tests 
wilcox.test(x = stat_lactobacteria$lactobacteria_wk1_abund, 
            y = stat_lactobacteria$lactobacteria_wk3_abund, 
            alternative = "less", paired = TRUE, 
            var.equal = TRUE) #p-value= .007803


#Violin Plot for Visualization
brmps_2x_lactobacteria_plot <- new_lactobacteria %>%
  filter(study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = lactobacteria)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = "Study Week",
       y = "Lacto Bacteria Relative Abundance",
       title = "Weekly Comparison of Abundances of Lacto Bacteria") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "brmps_2x_lactobacteria_plot.pdf",
          plot = brmps_2x_lactobacteria_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


brmps_2x_lactobacteria_plot
```

Confirming our hypothesis, the extremely low p-value (less than alpha = .01) provides significant evidence to reject the null hypothesis and accept the alternative. Thus, consuming the BRMPS supplement 2x daily did, on average, increase the mean number of Lacto Bacteria in participants from week 1 to week 3 of the study. We used a non-parametric (wilcox) significance test with equal variences because both shapiro tests gave significant evidence (indicated by their low p-values) that the subsets of data were not normal. Additionally, the histograms confirm that the data is vastly skewed to the right.