---
title: "Final Biology 201 Project"
author: "Seth Woodbury, Julie Heng, Maira Gajda, Karan Nayak"
date: "12/9/2019"
output: html_document
---
#Research Question:
Do any supplements cause a significant mean decrease in human gut pH from week 1 (pre-supplement) to week 3 (post-supplement)?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/bio201_project")
```

###Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq); packageVersion("phyloseq")
set.seed(7)
```

##Data Curation

###Import data
```{r}
#Import Data
shared_wkly <- read_delim(file = "raw_data/shared_wkly.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


biographical <- read_delim(file = "raw_data/biographical.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


ph_wkly <- read_delim(file = "raw_data/DB_v_0.08/pH_wkly.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) 


#Look at pH Data
ph_wkly
count(ph_wkly, Supplement_consumed) #to check number of data points per supplement
count(ph_wkly, Frequency) #to check frequency of supplement consumed
```
pH data was available for six supplements:
1. Banana (n=32)
2. BRMPS (n=365)
3. HiMaize+BRMPS (n=93)
4. LOODAT (n=35)
5. none (n=73)
6. transition_HiMaize (n=87)
and 2 with NA (remove)

###Weekly pH Data Curation
```{r}
#Rename to our standards and filter data
curated_ph_wkly <- ph_wkly %>%
  rename_all(tolower) %>%
  filter(quantity_compliant == "yes" | quantity_compliant == "none",
         use_data == "yes",
         study_week == "week1" | study_week == "week3",
         frequency != "NA") 
View(curated_ph_wkly)
write_delim(curated_ph_wkly, path = "curated_data/curated_ph_wkly.txt", delim = "\t")
```
Now we have a common standard for all of our data: snake case data, with compliant data from study weeks 1 and 3. This data set contains the general conditions for all further calculations. From here, we can further subset for specific conditions.


##Supplement-Specific Statistical Tests
Banana pH data was only given for one week, so we ommitted those values. The data we used came from the Fall 2018, Winter 2018, and Winter 2019 semesters.

####2x Daily BRMPS
```{r} 
#Subsetting data
brmps_wk1_2x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 
brmps_wk3_2x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 

#Normality & normality plots
shapiro.test(brmps_wk1_2x$ph_mean) #p-value = .3826 > 0.1, indicates normality
ggplot(brmps_wk1_2x, aes(x = ph_mean)) + geom_histogram() #unimodal, fairly normal
qqnorm(brmps_wk1_2x$ph_mean); qqline(brmps_wk1_2x$ph_mean) #did not follow the Normal Q-Q plot very accurately and the graph had a slight bimodal look

shapiro.test(brmps_wk3_2x$ph_mean) #p-value = .006713 < 0.1, indicates non-normality
ggplot(brmps_wk3_2x, aes(x = ph_mean)) + geom_histogram() #outlier on the left
qqnorm(brmps_wk3_2x$ph_mean); qqline(brmps_wk3_2x$ph_mean) #majority of points follows line generally


#Sample Size + Variance Assumptions
brmps_2x <- inner_join(x = brmps_wk1_2x, y = brmps_wk3_2x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(brmps_ph_mean_2x_wk1 = ph_mean.x,
         brmps_ph_mean_2x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))

str(brmps_2x) #n=27 < 30
var.test(x = brmps_2x$brmps_ph_mean_2x_wk1, 
         y = brmps_2x$brmps_ph_mean_2x_wk3, 
         alternative = "greater") #p-value = .9512 is not statistically significant, indicating equal variances
```
In calculating the conclusion, we decided to use a non-parametric t-test (Wilcox) because the assumptions for a t-test were not met (sample size was small, data was non-normal). 
```{r}
#Statistical Tests 
wilcox.test(x = brmps_2x$brmps_ph_mean_2x_wk1, 
            y = brmps_2x$brmps_ph_mean_2x_wk3, 
            alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.0002761
```
The extremely low p-value (less than alpha = .001) indicates that BRMPS taken 2x daily did significantly lower pH in participants from week 1 to week 3 sign. 


####2x Daily Transition Maize
```{r}
#Subsetting data
transition_wk1_2x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "transition_HiMaize", 
         frequency == "2xdaily") 
transition_wk3_2x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "transition_HiMaize", 
         frequency == "2xdaily") 

#Normality and normality plots
shapiro.test(transition_wk1_2x$ph_mean) #p-value = .8910 > 0.1, indicates normality
ggplot(transition_wk1_2x, aes(x = ph_mean)) + geom_histogram() #unimodal, normal, bell-shaped
qqnorm(transition_wk1_2x$ph_mean); qqline(transition_wk1_2x$ph_mean) #aligns with the line

shapiro.test(transition_wk3_2x$ph_mean) #p-value = .5083 > 0.1, indicates normality
ggplot(transition_wk3_2x, aes(x = ph_mean)) + geom_histogram() #fairly normal 
qqnorm(transition_wk3_2x$ph_mean); qqline(transition_wk3_2x$ph_mean) #aligns with the line


#Sample Size + Variance Assumptions
transition_2x <- inner_join(x = transition_wk1_2x, y = transition_wk3_2x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(transition_ph_mean_2x_wk1 = ph_mean.x,
         transition_ph_mean_2x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
str(transition_2x) #n=21 < 30
var.test(x = transition_2x$transition_ph_mean_2x_wk1, 
         y = transition_2x$transition_ph_mean_2x_wk3, 
         alternative = "greater") #p-value = .9395 > 0.1, indicates equal variances
```
Despite the smaller sample size, the assumptions indicating normality were met through the Shapiro test and associated histograms/QQ plots. Therefore we proceeded with a parametric t-test with equal variances.
```{r}
#Statistical Tests
t.test(x = transition_2x$transition_ph_mean_2x_wk1, y = transition_2x$transition_ph_mean_2x_wk3, 
       alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.007215 < 0.1, reject null hypothesis
```
The extremely low p-value (less than alpha = .001) indicates that Transition Maize taken 2x daily did significantly lower pH in participants from week 1 to week 3. 


####1x Daily BRMPS
```{r}
#Subsetting data
brmps_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
brmps_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 

#Normaily + normality plots
shapiro.test(brmps_wk1_1x$ph_mean) #p-value = 0.01669 < 0.1
ggplot(brmps_wk1_1x, aes(x = ph_mean)) + geom_histogram() #skewed left
qqnorm(brmps_wk1_1x$ph_mean); qqline(brmps_wk1_1x$ph_mean) #follows line closely

shapiro.test(brmps_wk3_1x$ph_mean) #p-value = .08154 < 0.1
ggplot(brmps_wk3_1x, aes(x = ph_mean)) + geom_histogram() #skewed left, unimodal
qqnorm(brmps_wk3_1x$ph_mean); qqline(brmps_wk3_1x$ph_mean) #follows line closely

#Sample Size + Variance Assumptions
brmps_1x <- inner_join(x = brmps_wk1_1x, y = brmps_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(brmps_ph_mean_1x_wk1 = ph_mean.x,
         brmps_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
str(brmps_1x) # n=62 > 30
var.test(x = brmps_1x$brmps_ph_mean_1x_wk1, 
         y = brmps_1x$brmps_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .6063 > 0.1, equal variances
```
We decided that a parametric test would be sufficient since the qq plots showed normal data and the histograms appeared to be approximately normal, even thought the sample size was not greater than 30. Additionally, we found the variances to be equal between the subsetted data sets.
```{r}
#Statistical Tests: 
t.test(x = brmps_1x$brmps_ph_mean_1x_wk1, y = brmps_1x$brmps_ph_mean_1x_wk3, 
       alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.1133
```
Although close to our alpha value of .10, we still could not reject our null hypothesis because the p-value was higher than the alpha value. Thus, BRMPS consumed 1x daily did not appear to significantly decrease the pH in participants' guts on average from week 1 to week 3. This being said, we must note that for a complex biological system with many variables, a 0.11 p-value is still noteworthy.


####1x Daily HiMaize+BRMPS
```{r}
#Subsetting data
himaize_brmps_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "HiMaize+BRMPS", 
         frequency == "1xdaily") 
himaize_brmps_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "HiMaize+BRMPS", 
         frequency == "1xdaily") 

#Normality and normality plots
shapiro.test(himaize_brmps_wk1_1x$ph_mean) #p-value = .2955
ggplot(himaize_brmps_wk1_1x, aes(x = ph_mean)) + geom_histogram() #looks unimodal, normal
qqnorm(himaize_brmps_wk1_1x$ph_mean); qqline(himaize_brmps_wk1_1x$ph_mean) #points follow line closely
shapiro.test(himaize_brmps_wk3_1x$ph_mean) #p-value = .05622
ggplot(himaize_brmps_wk3_1x, aes(x = ph_mean)) + geom_histogram() #unimodal, not even
qqnorm(himaize_brmps_wk3_1x$ph_mean); qqline(himaize_brmps_wk3_1x$ph_mean) #points follow line


#Sample Size + Variance Assumptions
himaize_brmps_1x <- inner_join(x = himaize_brmps_wk1_1x, y = himaize_brmps_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(himaize_brmps_ph_mean_1x_wk1 = ph_mean.x,
         himaize_brmps_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
str(himaize_brmps_1x) #n=23 < 30
var.test(x = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk1, 
         y = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .489 > 0.1, equal variances
```
We used a non-parametric test (wilcox), equal variances because week 3 had non-normal data according to the shapiro test and the graph for week 1 appeared to be bimodal in addition to a low sample size.
```{r}
#Statistical Tests 
wilcox.test(x = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk1, 
            y = himaize_brmps_1x$himaize_brmps_ph_mean_1x_wk3, 
            alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.05289
```
In this case, we found a significant p-value that was less than alpha = .10 providing us with sufficient evidence to reject the null hypothesis and accept the alternative conclusion that consuming HiMaize+BRMPS supplement 1x daily did decrease pH on average in individuals from week 1 to week 3. 


####1x Daily LOODAT
```{r}
#Subsetting data
loodat_wk1_1x <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "LOODAT", 
         frequency == "1xdaily") 
loodat_wk3_1x <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "LOODAT", 
         frequency == "1xdaily") 

#Normality and normality plots
shapiro.test(loodat_wk1_1x$ph_mean) #p-value = .1888
ggplot(loodat_wk1_1x, aes(x = ph_mean)) + geom_histogram() #normal, unimodal
qqnorm(loodat_wk1_1x$ph_mean); qqline(loodat_wk1_1x$ph_mean) #follow line closely
shapiro.test(loodat_wk3_1x$ph_mean) #p-value = .0213
ggplot(loodat_wk3_1x, aes(x = ph_mean)) + geom_histogram() #unimodal, with gaps in between (could constitute outliers)
qqnorm(loodat_wk3_1x$ph_mean); qqline(loodat_wk3_1x$ph_mean) #points tend to stray from line

#Sample Size + Variance Assumptions
loodat_1x <- inner_join(x = loodat_wk1_1x, y = loodat_wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(loodat_ph_mean_1x_wk1 = ph_mean.x,
         loodat_ph_mean_1x_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
str(loodat_1x) # n=15 < 30
var.test(x = loodat_1x$loodat_ph_mean_1x_wk1, 
         y = loodat_1x$loodat_ph_mean_1x_wk3, 
         alternative = "greater") #p-value = .3212 > 0.1, equal variances
```
We used a non-parametric test (wilcox) with equal variences because both subsetted data sets had bimodal data and there was a low sample size.
```{r}
#Statistical test: non-parametric
wilcox.test(x = loodat_1x$loodat_ph_mean_1x_wk1, 
            y = loodat_1x$loodat_ph_mean_1x_wk3, 
            alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.9024
```
The p-value here was far from significant, which would have required it to be lower than alpha = .10. We fail to reject the null hypothesis and therefore must conclude that consuming LOODAT 1x daily did not decrease the pH on average in participants' guts from week 1 to week 3. 


####No Supplement Consumed
```{r}
#Subsetting data
no_supplement_wk1 <- curated_ph_wkly %>%
  filter(study_week == "week1", 
         supplement_consumed == "none", 
         frequency == "-") 
no_supplement_wk3 <- curated_ph_wkly %>%
  filter(study_week == "week3", 
         supplement_consumed == "none", 
         frequency == "-") 

#Normality + Normality Plots
shapiro.test(no_supplement_wk1$ph_mean) #p-value = .1029
ggplot(no_supplement_wk1, aes(x = ph_mean)) + geom_histogram() #looks bimodal
qqnorm(no_supplement_wk1$ph_mean); qqline(no_supplement_wk1$ph_mean)
shapiro.test(no_supplement_wk3$ph_mean) #p-value = .03968
ggplot(no_supplement_wk3, aes(x = ph_mean)) + geom_histogram() #almost appears multimodal, but difficult to tell because of sample size limitations; several gaps, no clear bell shape of normality
qqnorm(no_supplement_wk3$ph_mean); qqline(no_supplement_wk3$ph_mean)

#Sample Size + Varience Assumptions
no_supplement <- inner_join(x = no_supplement_wk1, y = no_supplement_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(no_supplement_ph_mean_wk1 = ph_mean.x,
         no_supplement_ph_mean_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))
str(no_supplement) # n=24 < 30
var.test(x = no_supplement$no_supplement_ph_mean_wk1, 
         y = no_supplement$no_supplement_ph_mean_wk3, 
         alternative = "greater") #p-value = 0.2721
```
We surprisingly found the data sets to be bimodal when we graphed them and did we not have a large sample size leading us to use a non-parametric (wilcox) statistical test. In addition, we found the variances to be equal between the two subsetted data sets.
```{r}
#Statistical test: used non-parametric
wilcox.test(x = no_supplement$no_supplement_ph_mean_wk1, 
            y = no_supplement$no_supplement_ph_mean_wk3, 
            alternative = "greater", paired = TRUE, var.equal = TRUE) #p-value = 0.7781 > 0.1
```
In testing our control (no supplement) group, we hypothesized that there would be no significant result, and that is exactly what we got as the p-value was greater than alpha = .10. Thus we fail to reject the null hypothesis stating that there was no a decrease: on average for participants who did not consume a supplement did not see a change in pH from week 1 to week 3. 



##Finding the Mean Difference in pH between Week 3 and Week 1 for Each Supplement (Week 3 - Week 1)
The fact that some supplements caused a significant decrease in pH and others did not suggests that not all supplements produce the same results and that there might be an overarching supplement that caused the lowest drop in pH. Our goal here is to determine which supplement causes the largest mean difference in pH. We will only test the statistically significant data sets found above:
```{r}
#BRMPS 2x Daily Mean pH Difference
mean_diff_brmps_2x <- brmps_2x %>%
  select(participant_id, brmps_ph_mean_2x_wk1, brmps_ph_mean_2x_wk3) %>%
  mutate(ph_difference = brmps_ph_mean_2x_wk3 - brmps_ph_mean_2x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

View(mean_diff_brmps_2x)

#pH Difference = -0.3414815
```
This calculation tells us that participants who consumed BRMPS 2x daily had an average decrease of .34 in pH from week 1 to week 3.


```{r}
#Transition Maize 2x Daily Mean pH Difference
mean_diff_transition_2x <- transition_2x %>%
  select(participant_id, transition_ph_mean_2x_wk1, transition_ph_mean_2x_wk3) %>%
  mutate(ph_difference = transition_ph_mean_2x_wk3 - transition_ph_mean_2x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

View(mean_diff_transition_2x)

#pH Difference = -0.2895238
```
This calculation tells us that participants who consumed Transition Maize 2x daily had an average decrease of .29 in pH from week 1 to week 3.


```{r}
#HiMaize+BRMPS 1x Daily Mean pH Difference
mean_diff_himaize_brmps_1x <- himaize_brmps_1x %>%
  select(participant_id, himaize_brmps_ph_mean_1x_wk1, himaize_brmps_ph_mean_1x_wk3) %>%
  mutate(ph_difference = himaize_brmps_ph_mean_1x_wk3 - himaize_brmps_ph_mean_1x_wk1) %>%
  summarize(avg_ph_difference = mean(ph_difference))

View(mean_diff_himaize_brmps_1x)

#pH Difference = -0.2556522
```
This calculation tells us that participants who consumed HiMaize+BRMPS 1x daily had an average decrease of .26 in pH from week 1 to week 3.

As we can see, 2x Daily BRMPS causes the largest mean decrease in pH from weak 1 to weak 3, and thus we will focus on this supplement for the remainder of our statistical study.


#Species Diversity in BRMPS 2x Daily Individuals
```{r}
#Import Data
seq_var_table <- read_delim("raw_data/species_avg_shared.txt",
                            delim = "\t", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list()) %>%
  rename_all(tolower) %>%
  select(participant_id, study_week, semester, starts_with("Lactoc"), starts_with("Lactob"))
seq_var_table
```


```{r}
#Lactobacillus


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacillus <- inner_join(seq_var_table, brmps_2x_phylo) %>% #combine two phyloseq objects created above
  select(-starts_with("lactoco")) #get rid of lactococcus
lactobacillus


#Calculate Row Sums
lactobacillus_total <-  rowSums(lactobacillus[,4:39]) 
lactobacillus_total


#Mutate Row Sums into Previous Table
new_lactobacillus <- lactobacillus %>% 
  mutate(lactobacillus = lactobacillus_total) %>%
  select(participant_id, study_week, semester, lactobacillus, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacillus


#Correlation Data With Graph
wk_3_lactobacillus <- new_lactobacillus %>%
  filter(study_week == "week3")


wk_3_lactobacillus_plot <- wk_3_lactobacillus %>%
  ggplot(aes(x = ph_mean,
             y = lactobacillus)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactobacillus Bacteria") 
wk_3_lactobacillus_plot


#Correlation Test
lactobacillus_correlation <- wk_3_lactobacillus %>%
  lm(ph_mean ~ lactobacillus, data = .) #test relationship
summary(lactobacillus_correlation) #view results 
lactobacillus_correlation

#p-value = .4869
#R-squared = .01953
```
Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .4869) indicates that the null hypothesis is true meaning there is no correlation between pH and Lactobacillus genre bacteria. Additionally, the small r-squared value indicates that only about 1.9% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis which is opposite to what we hypothesized, as here the number of Lactobacillus bacteria increases as pH increases.


```{r}
#Lactococcus 


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactococcus <- inner_join(seq_var_table, brmps_2x_phylo) %>% #combine two phyloseq objects created above
  select(-starts_with("lactoba")) #get rid of lactococcus
lactococcus


#Calculate Row Sums
lactococcus_total <-  rowSums(lactococcus[,4:8]) 
lactococcus_total


#Mutate Row Sums into Previous Table
new_lactococcus <- lactococcus %>% 
  mutate(lactococcus = lactococcus_total) %>%
  select(participant_id, study_week, semester, lactococcus, ph_median, ph_mean, frequency, supplement_consumed)
new_lactococcus


#Correlation Data With Graph
wk_3_lactococcus <- new_lactococcus %>%
  filter(study_week == "week3")


wk_3_lactococcus_plot <- wk_3_lactococcus %>%
  ggplot(aes(x = ph_mean,
             y = lactococcus)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactococcus Bacteria") 
wk_3_lactococcus_plot


#Correlation Test
lactococcus_correlation <- wk_3_lactococcus %>%
  lm(ph_mean ~ lactococcus, data = .) #test relationship
summary(lactococcus_correlation) #view results 
lactococcus_correlation

#p-value = .3089
#R-squared = .04137
```
Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .3089) indicates that the null hypothesis is true meaning there is no correlation between pH and Lactococcus genre bacteria. Additionally, the small r-squared value indicates that only about 4.1% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis, which is opposite to what we hypothesized, as here the number of Lactococcus bacteria increases as pH increases.


We figured since everybody has a different microbiome, that some individuals may contain Lactococcus or just Lactobacillus genre bacteria and since they perform similar functions we thought it would be reasonable to combine them into one data set and see if there was a correlation between "Lacto" Bacteria and pH to compensate for different microbiome makeups.
```{r}
#"Lacto" Genre Bacteria 


#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacteria <- inner_join(seq_var_table, brmps_2x_phylo) 
lactobacteria


#Calculate Row Sums
lactobacteria_total <-  rowSums(lactobacteria[,4:44]) 
lactobacteria_total


#Mutate Row Sums into Previous Table
new_lactobacteria <- lactobacteria %>% 
  mutate(lactobacteria = lactobacteria_total) %>%
  select(participant_id, study_week, semester, lactobacteria, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacteria


#Correlation Data With Graph
wk_3_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week3")


wk_3_lactobacteria_plot <- wk_3_lactobacteria %>%
  ggplot(aes(x = ph_mean,
             y = lactobacteria)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 Mean pH") + 
  ylab("Relative Abundance of Lactate Producers") 
wk_3_lactobacteria_plot


#Correlation Test
lactobacteria_correlation <- wk_3_lactobacteria %>%
  lm(ph_mean ~ lactobacteria, data = .) #test relationship
summary(lactobacteria_correlation) #view results 
lactobacteria_correlation

#p-value = .4622
#R-squared = .02182
```
Just looking at the scatter plot we can see that there does not appear to be any correlation or pattern between the dots, as they appear to be randomly scattered. Our statistical analysis confirms this as the large p-value (p-value = .4622) indicates that the null hypothesis is true meaning there is no correlation between pH and "Lacto" bacteria. Additionally, the small r-squared value indicates that only about 2.2% of the data can be truly represented by the line of best fit, which is extremely small and inaccurate. If there were to be a correlation, it would be a positive one according to the statistical analysis, which is opposite to what we hypothesized, as here the number of "Lacto" bacteria increases as pH increases.


##Did the Number of Lactate Producers Rise?
Because our correlation tests did not find any significant results, we decided to look a little deeper to see if the supplement BRMPS taken 2x daily did at least increase the number of Lactate producers (Lactococcus and Lactobacillus bacteria). These Lactate producers may not be correlated with more acidic conditions, but they confer a variety of health benefits such as decreased gut permeability (inhibiting toxic bacterial biproducts from being absorbed) as well as disease resistance and increased immunity. Additionally, they produce acidic biproducts which may contribute to lower gut pH.
```{r}
#Import Individual Data
brmps_2x_phylo <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         frequency == "2xdaily")
brmps_2x_phylo


#Combine to Make New Data Sets
lactobacteria <- inner_join(seq_var_table, brmps_2x_phylo) 
lactobacteria


#Calculate Row Sums
lactobacteria_total <-  rowSums(lactobacteria[,4:44]) 
lactobacteria_total


#Mutate Row Sums into Previous Table
new_lactobacteria <- lactobacteria %>% 
  mutate(lactobacteria = lactobacteria_total) %>%
  select(participant_id, study_week, semester, lactobacteria, ph_median, ph_mean, frequency, supplement_consumed)
new_lactobacteria


#Weekly Data 
wk_1_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week1")


wk_3_lactobacteria <- new_lactobacteria %>%
  filter(study_week == "week3")


#Normality
shapiro.test(wk_1_lactobacteria$lactobacteria) 
shapiro.test(wk_3_lactobacteria$lactobacteria) 


#Normality Plots
ggplot(wk_1_lactobacteria, aes(x = lactobacteria)) + geom_histogram()
ggplot(wk_3_lactobacteria, aes(x = lactobacteria)) + geom_histogram()

qqnorm(wk_1_lactobacteria$lactobacteria); qqline(wk_1_lactobacteria$lactobacteria)
qqnorm(wk_3_lactobacteria$lactobacteria); qqline(wk_3_lactobacteria$lactobacteria)


#Final Joined Data
stat_lactobacteria <- inner_join(x = wk_1_lactobacteria, y = wk_3_lactobacteria,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed")) %>%
  rename(lactobacteria_wk1_abund = lactobacteria.x,
         lactobacteria_wk3_abund = lactobacteria.y) %>%
  select(-starts_with("study_week"))


#Sample Size + Varience Assumptions
str(stat_lactobacteria) # n=26
var.test(x = stat_lactobacteria$lactobacteria_wk1_abund, 
         y = stat_lactobacteria$lactobacteria_wk3_abund, 
         alternative = "greater") #p-value= 1


#Statistical Tests 
wilcox.test(x = stat_lactobacteria$lactobacteria_wk1_abund, 
            y = stat_lactobacteria$lactobacteria_wk3_abund, 
            alternative = "less", paired = TRUE, var.equal = TRUE)

#p-value= .007803


#Violin Plot for Visualization
brmps_2x_lactobacteria_plot <- new_lactobacteria %>%
  filter(study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = lactobacteria)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = "Study Week",
       y = "Lacto Bacteria Relative Abundance",
       title = "Weekly Comparison of Abundances of Lacto Bacteria") +
    theme(legend.position = "none")
brmps_2x_lactobacteria_plot


#Save Plot
save_plot(filename = "brmps_2x_lactobacteria_plot.pdf",
          plot = brmps_2x_lactobacteria_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)
```


#Visualizing Data
```{r}
#Transition HiMaize 2x Daily
transition_himaize_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "transition_HiMaize",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week)) +
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after 2x Transition HiMaize") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "transition_himaize_plot.pdf",
          plot = transition_himaize_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


transition_himaize_plot
```

```{r}
#BRMPS 1x Daily
brmps1_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS Supplement (1x daily)") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = brmps1_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


brmps1_plot
```


```{r}
#BRMPS 2x Daily
brmps2_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "BRMPS",
         study_week == "week1" | study_week == "week3",
         frequency == "2xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS Supplement (2x daily)") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = brmps2_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


brmps2_plot
```


```{r}
#HiMaize+BRMPS 1x Daily
himaize_brmps_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "HiMaize+BRMPS",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after HiMaize+BRMPS Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = himaize_brmps_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


himaize_brmps_plot
```

```{r}
#LOODAT 1x Daily
loodat_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "LOODAT",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after LOODAT Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = loodat_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


loodat_plot
```

```{r}
#No Supplement
no_supplement_plot <- curated_ph_wkly %>%
  filter(supplement_consumed == "none",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH without Supplement") +
    theme(legend.position = "none")


#Save Plot
save_plot(filename = "figures/no_supplement_plot.pdf",
          plot = no_supplement_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)


no_supplement_plot
```

