---
title: "Bio 201 Project Rough Draft - Julie"
author: "Julie Heng"
date: "11/18/2019" 
output: html_document
---

Goals: to determine which prebiotic has the greatest effect on pH levels (looking at most change). Once the prebiotics with the most change potential are identified, their effect on diversity, SCFAs, etc will be further studied to try to find specific pathways and mechanisms.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(phyloseq)
library(vegan)
set.seed(7)
```

#Curating data
We start by identifying usable data through data curation. 

```{r}
ph_data <- read_delim(file = "C:/Users/Jules/Documents/bio201_project/DB_v_0.08/ph_wkly.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA")) %>%
  rename_all(tolower) %>% 
  filter(quantity_compliant != "no") %>% 
  select(-use_data, -ph_median)

count(ph_data, supplement_consumed)
count(ph_data, frequency)
count(ph_data, semester)
```
Over 4 semesters (Fall 2017, Fall 2018, Winter 2018, Winter 2019), there were 5 supplements, banana, BRMPS, HiMaize+BRMPS, LOODAT, and transition_HiMaize. These are given either 1x or 2x daily. There is also a group tested without additional supplements. 


#Statistical tests
We are looking at how consuming these supplements affects pH. In this study, data points were collected before and after supplement consumption. We're using paired t-tests to evaluate whether weekly average pH values changed before and after consumption. A paired t-test will be conducted for every distinct different supplement consumed and frequency.

##Visualizing data
```{r}
#Banana x2 daily
banana_plot <- ph_data %>%
  filter(supplement_consumed == "Banana",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after Banana supplement") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/banana_plot.pdf",
          plot = banana_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#BRMPS x1 daily
brmps1_plot <- ph_data %>%
  filter(supplement_consumed == "Banana",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS supplement (once daily)") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps1_plot.pdf",
          plot = brmps1_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#BRMPS x2 daily
brmps2_plot <- ph_data %>%
  filter(supplement_consumed == "Banana",
         study_week == "week1" | study_week == "week3",
         frequency == "2xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after BRMPS supplement (twice daily)") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps2_plot.pdf",
          plot = brmps2_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#HiMaize+BRMPS x1 daily
himaize_brmps_plot <- ph_data %>%
  filter(supplement_consumed == "HiMaize+BRMPS",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after HiMaize+BRMPS supplement") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/himaize_brmps_plot.pdf",
          plot = himaize_brmps_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#LOODAT x1 daily
LOODAT_plot <- ph_data %>%
  filter(supplement_consumed == "LOODAT",
         study_week == "week1" | study_week == "week3",
         frequency == "1xdaily") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after LOODAT supplement") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/LOODAT_plot.pdf",
          plot = LOODAT_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#none
no_supplement_plot <- ph_data %>%
  filter(supplement_consumed == "none",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH without supplement") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/no_supplement_plot.pdf",
          plot = no_supplement_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)

#transition_HiMaize x2 daily
transition_himaize_plot <- ph_data %>%
  filter(supplement_consumed == "none",
         study_week == "week1" | study_week == "week3") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean)) + 
  geom_violin(aes(color = study_week)) +
  geom_jitter(aes(color = study_week))+
  labs(x = NULL,
       y = "Average weekly pH",
       title = "Change in pH after transition HiMaize supplement") +
    theme(legend.position = "none")
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/transition_himaize_plot.pdf",
          plot = transition_himaize_plot,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.1)
```

From these violin plots, we see pH decreases for the most part. Mostly, the lower bound decreases.

##Assumptions

Normality for paired t-tests. Visualize through histograms.

```{r}
ph_data_wk1 <- ph_data %>% 
  filter(study_week == "week1")
ph_data_wk3 <- ph_data %>% 
  filter(study_week == "week3")

banana_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before banana supplement") #appears normal, unimodal
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/banana_histogram1.pdf",
          plot = banana_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
banana_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after banana supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/banana_histogram3.pdf",
          plot = banana_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

brmps1_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before 1xdaily BRMPS supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps1_histogram1.pdf",
          plot = brmps1_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
brmps1_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after 1xdaily BRMPS supplement") #appears
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps1_histogram3.pdf",
          plot = brmps1_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

brmps2_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before 2xdaily BRMPS supplement") #=
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps2_histogram1.pdf",
          plot = brmps2_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
brmps2_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after 2xdaily BRMPS supplement") #
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/brmps2_histogram3.pdf",
          plot = brmps2_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

himaize_brmps_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before HiMaize+BRMPS supplement") #appears normal, unimodal
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/himaize_brmps_histogram1.pdf",
          plot = himaize_brmps_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
himaize_brmps_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after HiMaize+BRMPS supplement") #appears
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/himaize_brmps_histogram3.pdf",
          plot = himaize_brmps_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

LOODAT_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before LOODAT supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/LOODAT_histogram1.pdf",
          plot = LOODAT_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
LOODAT_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after LOODAT supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/LOODAT_histogram3.pdf",
          plot = LOODAT_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

no_supplement_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution with no supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/no_supplement_histogram1.pdf",
          plot = no_supplement_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
no_supplement_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after no supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/no_supplement_histogram3.pdf",
          plot = no_supplement_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)

transition_himaize_histogram1 <- ggplot(ph_data_wk1, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 1 pH distribution before Transition HiMaize supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/transition_himaize_histogram1.pdf",
          plot = transition_himaize_histogram1,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
transition_himaize_histogram3 <- ggplot(ph_data_wk3, aes(x=ph_mean)) +
  geom_histogram() +
  labs(title = "Histogram for week 3 pH distribution after Transition HiMaize supplement") #appears 
save_plot(filename = "C:/Users/Jules/Documents/bio201_project/figures/transition_himaize_histogram3.pdf",
          plot = transition_himaize_histogram3,
          nrow = 1, ncol = 1,
          base_aspect_ratio = 1.8)
```

